Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks

### 2024-12-23 _manuscript/_tex/index.tex
proofread Section 1-2

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Roger
D.
Peng. Email: \href{mailto:roger.peng@austin.utexas.edu}{\nolinkurl{roger.peng@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Texas, United States
\begin{abstract}
In data analysis, unexpected results often prompt researchers to revisit
their procedures to identify potential issues. While experienced
researchers can often quickly diagnose problems by checking a few key
assumptions, others may struggle to identify the root causes. These
checked assumptions, or expectations, are typically informal, difficult
to trace, and rarely discussed in publications. In this paper, we
formalize these informal assumptions by framing them as binary
\emph{analysis validation checks}. We then introduce a procedure to
quantify how violations of these checks may lead to unexpected results
in the analysis. The procedure relies on simulations of the original
data and evaluates both accuracy and redundancy. Accuracy is calculated
through a binary classification metric, while redundancy is measured
using mutual information. We demonstrate this approach with a toy
example based on fitness step count data and a generalized linear model
example examining the effect of particulate matter air pollution on
daily mortality.
\end{abstract}

@@

In data analysis, experienced researchers often rely on their prior
knowledge or domain expertise to quickly assess whether results align
with their expectations. When a result falls outside of this
expectation, it prompts the researchers to investigate backwards on the
data quality, the analysis steps, or the assumptions made during the
analysis process. This mental process of where to diagnose unexpected
outcomes is often difficult to trace and document in publications. As a
result, readers are typically presented with the final outcomes of the
analysis cycle where the results and expectations are aligned, achieved
either by refining the analysis or updating the expectations based on
statistical evidence \citep{grolemund_cognitive_2014}. These missing
pieces of information provide little guidance for diagnosing issues in
the analysis when the same methodology is applied to a new dataset that
produces different outcomes. Similarly, when researchers with different
background knowledge view results that they find to be unexpected, it
becomes unclear whether discrepancies arise from differing expectations
or from the use of statistical techniques.

One might gain insight into analysts' thought processes by speaking with
them directly or watching them work via screencast videos they produce,
such as, TidyTuesday screencast videos or think-aloud type studies
\citep[e.g.][]{gu2024data}. However, direct observation of analysis is
not scalable and may not always be feasible; creating educational
screencast videos requires significant effort from the researchers.
Ideally, there could be a way to make expectations about data analysis
explicit and accessible to others. Even better, if the encoding were
machine-readable, we could analyze these expectations and learn from the
analysis itself. For example, we could answer questions about whether
the checks also apply to other researchers analyzing new data in the
same context, whether they reflect common practices in the field, or
whether they are specific to the data or analysis at hand.

The externalization of the data analysis process is a practice that has
potential to improve the trustworthiness of analyses in general
\citep{yu2024veridical} and the trustworthiness of subsequent products,
such as machine learning models, that may be built on such analyses.
While publication of analysis code and data is now a common a
requirement for the sake of reproducibility
\citep{peng2011reproducible}, the published code alone is often
insufficient for understanding the thought process behind an analysis.
Code corresponding to published analyses often reflect the final
decisions made about analysis and do not reveal the decisions or
assumptions made about the data processing. Thus, a reader looking at
published code can often be left with many questions about why certain
choices were made. Developing an approach to reveal some of this process
without requiring a reader to essentially reconstruct the analysis
process from the raw data would provide an improved basis for trusting
an analysis result \citep{peng2021reproducible}.

In this paper, we conceptualize these internal expectations and
assumptions as \emph{analysis validation checks}, which allows us to
examine the assumptions made during an analysis and to diagnose
unexpected outcomes. We then introduce a procedure that provides a
quantitative measure of how violations in a tree of analysis checks,
derived from individual checks, will lead to an unexpected result. The
procedure, based on simulations of the original data, calculates the
accuracy and redundancy of the analysis checks. Accuracy is determined
using binary classification metrics, precision and recall, from a logic
regression fit \citep{ruczinski_logic_2003}, while redundancy is
measured using mutual information. The proposed workflow offers a
numerical guarantee that the analysis will produce the expected results,
assuming the assumptions about the data generating mechanism hold.

The rest of the paper is organized as follows:
Section~\ref{sec-lit-review} reviews the concepts of diagnosing
unexpected outcomes and general data quality checks.
Section~\ref{sec-plan} introduces the concept of analysis validation
checks, illustrated with a toy example based on fitness step count data.
Section~\ref{sec-method} describes the procedure that quantifies how
analysis validation checks combined using logical operators can predict
unexpected outcomes in an analysis. Section~\ref{sec-pm10-mortality}
applies this procedure to a larger example that estimates the effect of
particulate matter air pollution on daily mortality.
Section~\ref{sec-discussion} summarises teh paper and discusses a few
key considerations.

@@

A substantial body of literature has addressed the definition of data
quality and has developed frameworks that include dimensions,
attributes, and measures to evaluate and improve data quality
\citep{8642813, batini2009methodologies, cai2015challenges, wang1996beyond, 6204995, woodall2014classification}.
These frameworks are often used in information systems and database
management and support business decision-making in various industries.
For example, such data validation systems are applied in the Google
Tensorflow framework \citep{polyzotis2019data} and Amazon
\citep{schelter2018automating}. For research purposes, high-quality data
ensures the credibility of scientific findings and supports
reproducibility and re-usability in future studies. With the growing
prevalence of open data in scientific research, the consumers or users
of the data typically are no longer the producers or collectors of the
data who would have the full knowledge of data in hand, prompting more
interest towards data quality checks in the data analysis process. In R,
there are some packages, like \texttt{skimr} \citep{skimr} and
\texttt{dataMaid} \citep{dataMaid}, that provide basic data screening
and reporting tools, while another class of packages,
e.g.~\texttt{assertr} \citep{assertr}, \texttt{validate}
\citep{validate}, and \texttt{pointblank} \citep{pointblank} focuses on
providing data validation tools, allowing users to define customized
data quality checks based on the applications.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Roger
D.
Peng. Email: \href{mailto:roger.peng@austin.utexas.edu}{\nolinkurl{roger.peng@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Texas, United States
\begin{abstract}
In data analysis, unexpected results often prompt researchers to revisit
their procedures to identify potential issues. While some researchers
may struggle to identify the root causes, experienced researchers can
often quickly diagnose problems by checking a few key assumptions. These
checked assumptions, or expectations, are typically informal, difficult
to trace, and rarely discussed in publications. In this paper, we
introduce the term \emph{analysis validation checks} to formalize and
externalize these informal assumptions. We then introduce a procedure to
identify a subset of checks that best predict the occurrence of
unexpected outcomes, based on simulations of the original data. The
checks are evaluated in terms of accuracy, determined by binary
classification metrics, and redundancy, which measures the shared
information among checks using mutual information. We demonstrate this
approach with a toy example using step count data and a generalized
linear model example examining the effect of particulate matter air
pollution on daily mortality.
\end{abstract}

@@

In data analysis, researchers often rely on their prior knowledge or
domain expertise to form expectations and assess whether results align
with them. When results deviate from these expectations, experienced
researchers check assumptions made about the data and during the
analysis, but these assumptions checked are often difficult to trace and
are seldom documented in publications. While publication of analysis
code and data is now a common a requirement for the sake of
reproducibility \citep{peng2011reproducible}, the published code alone
is often insufficient for understanding the thought process behind an
analysis. Code corresponding to published analyses often reflect the
final decisions made about analysis and do not reveal the decisions or
assumptions made about the data during the analysis. Thus, a reader
looking at published code can often be left with many questions about
why certain choices were made.

We might gain insight into analysts' thought processes by speaking with
them directly or watching them work via screencast videos they produce,
such as, TidyTuesday screencast videos or think-aloud type studies
\citep[e.g.][]{gu2024data}. However, direct observation of analysis is
not scalable and may not always be feasible; creating educational
screencast videos requires significant effort from the analysts.
Ideally, there could be a way to explicitly communciate these throught
processes in the data analysis to others. Even better, if these
expectations were machine-readable, we could analyze them and learn from
the analysis itself. For example, we could answer questions about
whether the checks also apply to other researchers analyzing new data in
the same context, whether they reflect common practices in the field, or
whether they are specific to the data or analysis at hand. Externalizing
these thought processes is a practice that has potential to improve the
trustworthiness of the analyses \citep{yu2024veridical} in general and
the trustworthiness of subsequent products, such as machine learning
models, that may be built on such analyses. Developing an approach to
reveal some of this process without requiring a reader to essentially
reconstruct the analysis process from the raw data would provide an
improved basis for trusting an analysis result
\citep{peng2021reproducible}.

In this paper, we conceptualize these inexplicit expectations as
\emph{analysis validation checks}, which allows us to examine the
assumptions made about the data during an analysis. We then introduce a
procedure to identify the subset of checks that best predict the
occurrence of unexpected outcomes. The procedure, based on simulations
of the original data, compares the accuracy and redundancy of different
combinations of analysis validation checks. Accuracy is determined using
binary classification metrics (precision and recall) from a logic
regression model \citep{ruczinski_logic_2003}, while redundancy is
measured using mutual information. The proposed workflow offers a
numerical guarantee that the analysis will produce the expected results,
assuming the assumptions about the data generating mechanism hold.

The rest of the paper is organized as follows:
Section~\ref{sec-lit-review} reviews the concepts of diagnosing
unexpected outcomes and general data quality checks.
Section~\ref{sec-plan} introduces the concept of analysis validation
checks, illustrated with a toy example based on step count data.
Section~\ref{sec-method} describes the procedure that selects the best
subset of checks for understanding the occurrence of unexpected
outcomes. Section~\ref{sec-pm10-mortality} applies this procedure to a
larger example that estimates the effect of particulate matter air
pollution on daily mortality. Section~\ref{sec-discussion} summarises
the paper and discusses a few key considerations.

@@

The concept of data quality has been studied in the literature, with
earlier work focusing on defining frameworks -- such as dimensions,
attributes, and measures -- to improve data quality in database and
information systems
\citep{wang1996beyond, batini2009methodologies, 6204995, woodall2014classification, cai2015challenges, 8642813}.
More recently, data validation has been incorporated into frameworks
like Google TensorFlow \citep{polyzotis2019data} to ensure the quality
of data for training machine learning models, as well as for supporting
business decision-making \citep{schelter2018automating}. With the
increasing availability of open data in scientific research, the users
of the data are often not the original data collector and may not be
aware of all the details or nuances of the data. This encourages
researchers to conduct data quality checks before beginning the analytic
process, helping to avoid unexpected discoveries later on. In R,
packages like \texttt{skimr} \citep{skimr} and \texttt{dataMaid}
\citep{dataMaid} provide basic data screening and data quality reports,
while packages such as \texttt{assertr} \citep{assertr},
\texttt{validate} \citep{validate}, and \texttt{pointblank}
\citep{pointblank} focuse on providing infustructures that allow users
to define customized data quality checks.

### 2024-12-23 index.qmd
proofread Section 1-2

---
title: "Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks"
abstract: "In data analysis, unexpected results often prompt researchers to revisit their procedures to identify potential issues. While experienced researchers can often quickly diagnose problems by checking a few key assumptions, others may struggle to identify the root causes. These checked assumptions, or expectations, are typically informal, difficult to trace, and rarely discussed in publications. In this paper, we formalize these informal assumptions by framing them as binary *analysis validation checks*. We then introduce a procedure to quantify how violations of these checks may lead to unexpected results in the analysis. The procedure relies on simulations of the original data and evaluates both accuracy and redundancy. Accuracy is calculated through a binary classification metric, while redundancy is measured using mutual information. We demonstrate this approach with a toy example based on fitness step count data and a generalized linear model example examining the effect of particulate matter air pollution on daily mortality. "
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Texas
          country: United States
          postal-code: 78705
    attributes:
        corresponding: true
  - name: Roger D. Peng
    email: roger.peng@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Texas
          country: United States
          postal-code: 78705
cite-method: natbib
bibliography: references.bib
format:
 tandf-pdf:
   keep-tex: true
   fontsize: 12pt
   linestretch: 1.15
   notebook-links: false
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
editor_options: 
  chunk_output_type: console
---

@@

In data analysis, experienced researchers often rely on their prior knowledge or domain expertise to quickly assess whether results align with their expectations. When a result falls outside of this expectation, it prompts the researchers to investigate backwards on the data quality, the analysis steps, or the assumptions made during the analysis process. This mental process of where to diagnose unexpected outcomes is often difficult to trace and document in publications. As a result, readers are typically presented with the final outcomes of the analysis cycle where the results and expectations are aligned, achieved either by refining the analysis or updating the expectations based on statistical evidence [@grolemund_cognitive_2014]. These missing pieces of information provide little guidance for diagnosing issues in the analysis when the same methodology is applied to a new dataset that produces different outcomes. Similarly, when researchers with different background knowledge view results that they find to be unexpected, it becomes unclear whether discrepancies arise from differing expectations or from the use of statistical techniques.

One might gain insight into analysts' thought processes by speaking with them directly or watching them work via screencast videos they produce, such as, TidyTuesday screencast videos or think-aloud type studies [e.g. @gu2024data]. However, direct observation of analysis is not scalable and may not always be feasible; creating educational screencast videos requires significant effort from the researchers. Ideally, there could be a way to make expectations about data analysis explicit and accessible to others. Even better, if the encoding were machine-readable, we could analyze these expectations and learn from the analysis itself. For example, we could answer questions about whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect common practices in the field, or whether they are specific to the data or analysis at hand.

The externalization of the data analysis process is a practice that has potential to improve the trustworthiness of analyses in general [@yu2024veridical] and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses. While publication of analysis code and data is now a common a requirement for the sake of reproducibility [@peng2011reproducible], the published code alone is often insufficient for understanding the thought process behind an analysis. Code corresponding to published analyses often reflect the final decisions made about analysis and do not reveal the decisions or assumptions made about the data processing. Thus, a reader looking at published code can often be left with many questions about why certain choices were made. Developing an approach to reveal some of this process without requiring a reader to essentially reconstruct the analysis process from the raw data would provide an improved basis for trusting an analysis result [@peng2021reproducible].

In this paper, we conceptualize these internal expectations and assumptions as *analysis validation checks*, which allows us to examine the assumptions made during an analysis and to diagnose unexpected outcomes. We then introduce a procedure that provides a quantitative measure of how violations in a tree of analysis checks, derived from individual checks, will lead to an unexpected result. The procedure, based on simulations of the original data, calculates the accuracy and redundancy of the analysis checks. Accuracy is determined using binary classification metrics, precision and recall, from a logic regression fit [@ruczinski_logic_2003], while redundancy is measured using mutual information. The proposed workflow offers a numerical guarantee that the analysis will produce the expected results, assuming the assumptions about the data generating mechanism hold.

The rest of the paper is organized as follows: @sec-lit-review reviews the concepts of diagnosing unexpected outcomes and general data quality checks. @sec-plan introduces the concept of analysis validation checks, illustrated with a toy example based on fitness step count data. @sec-method describes the procedure that quantifies how analysis validation checks combined using logical operators can predict unexpected outcomes in an analysis. @sec-pm10-mortality applies this procedure to a larger example that estimates the effect of particulate matter air pollution on daily mortality. @sec-discussion summarises teh paper and discusses a few key considerations.

@@

A substantial body of literature has addressed the definition of data quality and has developed frameworks that include dimensions, attributes, and measures to evaluate and improve data quality [@8642813; @batini2009methodologies; @cai2015challenges; @wang1996beyond; @6204995; @woodall2014classification]. These frameworks are often used in  information systems and database management and support business decision-making in various industries. For example, such data validation systems are applied in the Google Tensorflow framework [@polyzotis2019data] and Amazon [@schelter2018automating]. For research purposes, high-quality data ensures the credibility of scientific findings and supports reproducibility and re-usability in future studies. With the growing prevalence of open data in scientific research, the consumers or users of the data typically are no longer the producers or collectors of the data who would have the full knowledge of data in hand, prompting more interest towards data quality checks in the data analysis process. In R, there are some packages, like `skimr` [@skimr] and `dataMaid` [@dataMaid], that provide basic data screening and reporting tools, while another class of packages, e.g. `assertr` [@assertr], `validate` [@validate], and `pointblank` [@pointblank] focuses on providing data validation tools, allowing users to define customized data quality checks based on the applications.

>>>>

---
title: "Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks"
abstract: "In data analysis, unexpected results often prompt researchers to revisit their procedures to identify potential issues. While some researchers may struggle to identify the root causes, experienced researchers can often quickly diagnose problems by checking a few key assumptions. These checked assumptions, or expectations, are typically informal, difficult to trace, and rarely discussed in publications. In this paper, we introduce the term *analysis validation checks* to formalize and externalize these informal assumptions. We then introduce a procedure to identify a subset of checks that best predict the occurrence of unexpected outcomes, based on simulations of the original data. The checks are evaluated in terms of accuracy, determined by binary classification metrics, and redundancy, which measures the shared information among checks using mutual information. We demonstrate this approach with a toy example using step count data and a generalized linear model example examining the effect of particulate matter air pollution on daily mortality. "
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Texas
          country: United States
          postal-code: 78705
    attributes:
        corresponding: true
  - name: Roger D. Peng
    email: roger.peng@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Texas
          country: United States
          postal-code: 78705
cite-method: natbib
bibliography: references.bib
format:
 tandf-pdf:
   keep-tex: true
   fontsize: 12pt
   linestretch: 1.15
   notebook-links: false
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
editor_options: 
  chunk_output_type: console
---

@@

In data analysis, researchers often rely on their prior knowledge or domain expertise to form expectations and assess whether results align with them. When results deviate from these expectations, experienced researchers check assumptions made about the data and during the analysis, but these assumptions checked are often difficult to trace and are seldom documented in publications. While publication of analysis code and data is now a common a requirement for the sake of reproducibility [@peng2011reproducible], the published code alone is often insufficient for understanding the thought process behind an analysis. Code corresponding to published analyses often reflect the final decisions made about analysis and do not reveal the decisions or assumptions made about the data during the analysis. Thus, a reader looking at published code can often be left with many questions about why certain choices were made.

We might gain insight into analysts' thought processes by speaking with them directly or watching them work via screencast videos they produce, such as, TidyTuesday screencast videos or think-aloud type studies [e.g. @gu2024data]. However, direct observation of analysis is not scalable and may not always be feasible; creating educational screencast videos requires significant effort from the analysts. Ideally, there could be a way to explicitly communciate these throught processes in the data analysis to others. Even better, if these expectations were machine-readable, we could analyze them and learn from the analysis itself. For example, we could answer questions about whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect common practices in the field, or whether they are specific to the data or analysis at hand. Externalizing these thought processes is a practice that has potential to improve the trustworthiness of the analyses [@yu2024veridical] in general and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses. Developing an approach to reveal some of this process without requiring a reader to essentially reconstruct the analysis process from the raw data would provide an improved basis for trusting an analysis result [@peng2021reproducible].

In this paper, we conceptualize these inexplicit expectations as *analysis validation checks*, which allows us to examine the assumptions made about the data during an analysis. We then introduce a procedure to identify the subset of checks that best predict the occurrence of unexpected outcomes. The procedure, based on simulations of the original data, compares the accuracy and redundancy of different combinations of analysis validation checks. Accuracy is determined using binary classification metrics (precision and recall) from a logic regression model [@ruczinski_logic_2003], while redundancy is measured using mutual information. The proposed workflow offers a numerical guarantee that the analysis will produce the expected results, assuming the assumptions about the data generating mechanism hold.

The rest of the paper is organized as follows: @sec-lit-review reviews the concepts of diagnosing unexpected outcomes and general data quality checks. @sec-plan introduces the concept of analysis validation checks, illustrated with a toy example based on step count data. @sec-method describes the procedure that selects the best subset of checks for understanding the occurrence of unexpected outcomes. @sec-pm10-mortality applies this procedure to a larger example that estimates the effect of particulate matter air pollution on daily mortality. @sec-discussion summarises the paper and discusses a few key considerations.

@@

The concept of data quality has been studied in the literature, with earlier work focusing on defining frameworks -- such as dimensions, attributes, and measures -- to improve data quality in database and information systems [@wang1996beyond; @batini2009methodologies; @6204995; @woodall2014classification; @cai2015challenges; @8642813]. More recently, data validation has been incorporated into frameworks like Google TensorFlow [@polyzotis2019data] to ensure the quality of data for training machine learning models, as well as for supporting business decision-making [@schelter2018automating]. With the increasing availability of open data in scientific research, the users of the data are often not the original data collector and may not be aware of all the details or nuances of the data. This encourages researchers to conduct data quality checks before beginning the analytic process, helping to avoid unexpected discoveries later on. In R, packages like `skimr` [@skimr] and `dataMaid` [@dataMaid] provide basic data screening and data quality reports, while packages such as  `assertr` [@assertr], `validate` [@validate], and `pointblank` [@pointblank] focuse on providing infustructures that allow users to define customized data quality checks.
